---
title: "Commonality Inference: Paper Analyses"
author: "https://github.com/SocialInteractionLab/shared-reality"
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
engine: knitr
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    message: false
    warning: false
---

This document reproduces all behavioral analyses and figures from the paper.

1. **Setup** - Imports and data loading
2. **Manipulation Check** - Stance recognition (Paper p.6)
3. **Figure 2** - Generalization gradient with condition overlay + SR-G prediction (Paper p.8-9)
4. **Domain Analysis** - Figure 3 + 42 domain-specific models (Paper p.9)
5. **Sample Description** - Methods section

## 1. Setup

```{r}
#| label: setup
#| include: false
library(reticulate)
use_python(Sys.which("python3"), required = TRUE)
library(lme4)
library(lmerTest)
library(dplyr)
library(tidyr)
library(purrr)
```

```{python}
import sys
from pathlib import Path
import pandas as pd
import numpy as np
import seaborn as sns
import matplotlib.pyplot as plt
from scipy.stats import pearsonr, linregress
from scipy import stats
from statsmodels.stats.multitest import multipletests
import polars as pl
from polars import col, lit

plt.rcParams['font.family'] = 'Helvetica Neue'
sns.set_style('whitegrid')
SAVE_FIGURES = True
```

```{python}
# Find project root
base_dir = Path.cwd().resolve()
while base_dir.name and not (base_dir / 'data').exists():
    base_dir = base_dir.parent
if not (base_dir / 'data').exists():
    raise ValueError('Could not find project root with data directory')

data_dir = base_dir / 'data'
figures_dir = base_dir / 'outputs' / 'figures'
figures_dir.mkdir(parents=True, exist_ok=True)

# Stance palette: opposing (blue), shared (pink)
stance_palette = ['#648FFF', '#DC267F']
print(f'Data directory: {data_dir}')
```

### Load and preprocess data

```{python}
# Load pre-merged behavioral data
df = pl.read_csv(data_dir / 'responses.csv')
df = df.filter(~col('matchedTolerance').is_null())

# Derived variables
df = df.with_columns([
    pl.when(col('matchedTolerance') <= 1).then(lit('shared')).otherwise(lit('opposing')).alias('stance'),
    pl.when(col('matchedQuestion') == col('preChatQuestion')).then(lit('sameQ'))
     .when(col('matchedDomain') == col('preChatDomain')).then(lit('sameD'))
     .otherwise(lit('diffD')).alias('category'),
    (col('postChatResponse') - col('preChatResponse')).abs().alias('distance')
]).rename({'groupId': 'dyadId'})

# Convert to pandas with numeric contrasts
df_pd = df.to_pandas()
df_pd['stance_num'] = df_pd['stance'].map({'opposing': -1, 'shared': 1})
df_pd['category_num'] = df_pd['category'].map({'sameQ': 1, 'sameD': 0, 'diffD': -1}).astype(float)
df_pd['experiment_num'] = df_pd['experiment'].map({'no-chat': -1, 'chat': 1})
df_pd['distance_num'] = df_pd['distance'].map({0: 2, 1: 1, 2: 0, 3: -1, 4: -2})

print(f'Loaded {len(df_pd)} observations from {df_pd["pid"].nunique()} participants')
```

```{python}
#| output: false
# Save main data for R (once)
df_pd.to_csv('/tmp/main_df.csv', index=False)
```

## 2. Manipulation Check

Participants should accurately perceive agreement/disagreement on the focal topic. Paper reports the 2×2 breakdown: "judging it as shared far more often when they actually agreed (79% observation-only, 89% chat) than when they disagreed (19% observation-only, 38% chat)."

```{python}
# Filter to observed/focal question only
obs_df = df_pd[df_pd['question'] == df_pd['matchedIdx']]

print('=== Manipulation Check: P(judged shared) on FOCAL question ===\n')

# 2x2 table
print('                        Actual Stance')
print('Interaction        Shared          Opposing')
print('-' * 50)

manip_results = []
for exp in ['no-chat', 'chat']:
    for stance in ['shared', 'opposing']:
        cell = obs_df[(obs_df['experiment'] == exp) & (obs_df['stance'] == stance)]
        n = len(cell)
        p_shared = cell['predictShared'].mean()
        manip_results.append({
            'experiment': exp,
            'stance': stance,
            'n': n,
            'p_shared': p_shared
        })

manip_df = pd.DataFrame(manip_results)

for exp in ['no-chat', 'chat']:
    exp_data = manip_df[manip_df['experiment'] == exp]
    shared = exp_data[exp_data['stance'] == 'shared'].iloc[0]
    opposing = exp_data[exp_data['stance'] == 'opposing'].iloc[0]
    label = "No-Chat" if exp == 'no-chat' else "Chat"
    print(f"{label:12}   {shared['p_shared']*100:5.1f}% (n={int(shared['n'])})   {opposing['p_shared']*100:5.1f}% (n={int(opposing['n'])})")

print('-' * 50)

# Discrimination by condition
print('\n=== Discrimination (Shared - Opposing) ===\n')
for exp in ['no-chat', 'chat']:
    exp_data = manip_df[manip_df['experiment'] == exp]
    p_shared = exp_data[exp_data['stance'] == 'shared']['p_shared'].values[0]
    p_opposing = exp_data[exp_data['stance'] == 'opposing']['p_shared'].values[0]
    discrim = p_shared - p_opposing
    label = "No-Chat" if exp == 'no-chat' else "Chat"
    print(f"{label:12}: {discrim*100:+.1f} percentage points")

print('\n→ Both conditions show strong discrimination')
print('→ Chat elevates commonality judgments in BOTH cells')
print('  (even when partners disagreed, conversation led to perceiving more common ground)')
```

## 3. Figure 2: Generalization Gradient and SR-G Prediction

Paper reports:
- Stance × question type interaction β = 0.54, 95% CI [0.48, 0.59] (p.8-9)
- Commonality → SR-G β = 1.84, CI [1.60, 2.08]; three-way interaction β = −0.47, CI [−0.71, −0.23] (p.8)

```{python}
#| fig-cap: "Figure 2: (A) Generalization gradient with condition overlay, (B) SR-G prediction"
#| fig-width: 10
#| fig-height: 4

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from matplotlib.patches import Patch
from scipy import stats

# Reload data and settings
df_pd = pd.read_csv('/tmp/main_df.csv', low_memory=False)
from pathlib import Path
base_dir = Path.cwd().resolve()
while base_dir.name and not (base_dir / 'data').exists():
    base_dir = base_dir.parent
figures_dir = base_dir / 'outputs' / 'figures'
SAVE_FIGURES = True

# Colors and style
BLUE, PINK = '#648FFF', '#DC267F'
plt.rcParams.update({
    'font.family': 'Helvetica Neue',
    'font.size': 12,
    'axes.labelsize': 14,
    'axes.titlesize': 14,
    'xtick.labelsize': 12,
    'ytick.labelsize': 12,
    'legend.fontsize': 11,
    'axes.spines.top': False,
    'axes.spines.right': False,
})

def bootstrap_ci(data, n_boot=1000, ci=95, seed=42):
    """Compute bootstrap confidence interval for the mean."""
    np.random.seed(seed)
    boot_means = [np.nanmean(np.random.choice(data, size=len(data), replace=True))
                  for _ in range(n_boot)]
    return np.percentile(boot_means, (100 - ci) / 2), np.percentile(boot_means, 100 - (100 - ci) / 2)

fig, axes = plt.subplots(1, 2, figsize=(10, 4))

# =========================================================================
# Panel A: Pooled gradient with condition overlay
# =========================================================================
ax = axes[0]

category_order = ['sameQ', 'sameD', 'diffD']
category_labels = ['Observed', 'Same Domain', 'Different Domain']
stance_order = ['opposing', 'shared']
colors = [BLUE, PINK]

bar_width = 0.35
x_positions = np.arange(len(category_order))
jitter = 0.05
dot_size = 70

# Compute pooled means and CIs
pooled_means, pooled_cis = {}, {}
for stance in stance_order:
    for cat in category_order:
        subset = df_pd[(df_pd['stance'] == stance) & (df_pd['category'] == cat)]
        pooled_means[(stance, cat)] = subset['predictShared'].mean()
        pooled_cis[(stance, cat)] = bootstrap_ci(subset['predictShared'].values)

# Compute condition-specific means
condition_means = {}
for exp in ['no-chat', 'chat']:
    for stance in stance_order:
        for cat in category_order:
            subset = df_pd[(df_pd['experiment'] == exp) &
                          (df_pd['stance'] == stance) &
                          (df_pd['category'] == cat)]
            condition_means[(exp, stance, cat)] = subset['predictShared'].mean()

# Plot bars
for i, (stance, color) in enumerate(zip(stance_order, colors)):
    offset = (i - 0.5) * bar_width
    means = [pooled_means[(stance, cat)] for cat in category_order]
    lowers = [pooled_means[(stance, cat)] - pooled_cis[(stance, cat)][0] for cat in category_order]
    uppers = [pooled_cis[(stance, cat)][1] - pooled_means[(stance, cat)] for cat in category_order]
    ax.bar(x_positions + offset, means, bar_width, color=color, alpha=0.8,
           yerr=[lowers, uppers], capsize=0, error_kw={'linewidth': 1.5})

# Overlay condition dots (no connecting lines)
from matplotlib.lines import Line2D

for i, (stance, color) in enumerate(zip(stance_order, colors)):
    offset = (i - 0.5) * bar_width
    for j, cat in enumerate(category_order):
        x_base = x_positions[j] + offset
        nochat_mean = condition_means[('no-chat', stance, cat)]
        chat_mean = condition_means[('chat', stance, cat)]
        x_nochat, x_chat = x_base - jitter, x_base + jitter

        # Open circle for no-chat, filled circle for chat
        ax.scatter(x_nochat, nochat_mean, s=dot_size, facecolor='white', edgecolor=color, linewidth=2, zorder=3)
        ax.scatter(x_chat, chat_mean, s=dot_size, facecolor=color, edgecolor='white', linewidth=1.5, zorder=3)

ax.set_xlabel('')
ax.set_ylabel('Expected Commonality')
ax.set_xticks(x_positions)
ax.set_xticklabels(category_labels)
ax.set_ylim([0, 1])

# Legend: Shared first, then Opposing, then dot types
legend_elements = [
    Patch(facecolor=PINK, alpha=0.8, label='Shared'),
    Patch(facecolor=BLUE, alpha=0.8, label='Opposing'),
    Line2D([0], [0], marker='o', color='w', markerfacecolor='black',
           markeredgecolor='black', markersize=8, linestyle='None', label='Chat'),
    Line2D([0], [0], marker='o', color='w', markerfacecolor='white',
           markeredgecolor='black', markeredgewidth=1.5, markersize=8,
           linestyle='None', label='No chat'),
]
ax.legend(handles=legend_elements, loc='upper right')

ax.text(-0.12, 1.05, 'A', transform=ax.transAxes, fontsize=14, fontweight='bold')

# =========================================================================
# Panel B: SR-G prediction (pooled across interaction types)
# =========================================================================
ax = axes[1]

inf_df = df_pd[df_pd['question'] != df_pd['matchedIdx']]
srg_df = inf_df.groupby(['stance', 'pid', 'matchedDomain']).agg({
    'predictShared': 'mean', 'srgiResponse': 'mean'
}).reset_index()
# Also save version with experiment for R model
srg_df_full = inf_df.groupby(['experiment', 'stance', 'pid', 'matchedDomain']).agg({
    'predictShared': 'mean', 'srgiResponse': 'mean',
    'experiment_num': 'first', 'stance_num': 'first'
}).reset_index()
srg_df_full.to_csv('/tmp/srg_df.csv', index=False)

for stance, color in zip(stance_order, colors):
    subset = srg_df[srg_df['stance'] == stance]
    ax.scatter(subset['predictShared'], subset['srgiResponse'], c=color, alpha=0.4, s=30)

    x, y = subset['predictShared'].values, subset['srgiResponse'].values
    mask = ~(np.isnan(x) | np.isnan(y))
    x, y = x[mask], y[mask]

    if len(x) > 2:
        slope, intercept, r, p, se = stats.linregress(x, y)
        x_line = np.linspace(x.min(), x.max(), 100)
        y_line = slope * x_line + intercept
        n, x_mean, ss_x = len(x), np.mean(x), np.sum((x - np.mean(x)) ** 2)
        se_y = np.sqrt(np.sum((y - (slope * x + intercept)) ** 2) / (n - 2))
        t_val = stats.t.ppf(0.975, n - 2)
        se_line = se_y * np.sqrt(1/n + (x_line - x_mean) ** 2 / ss_x)
        ax.plot(x_line, y_line, color=color, linewidth=2)
        ax.fill_between(x_line, y_line - t_val * se_line, y_line + t_val * se_line, color=color, alpha=0.2)

ax.set_xlabel('Expected Commonality')
ax.set_ylabel('SR-G Score')
ax.set_xlim([0, 1])
ax.text(-0.12, 1.05, 'B', transform=ax.transAxes, fontsize=14, fontweight='bold')

plt.tight_layout()
if SAVE_FIGURES:
    plt.savefig(figures_dir / 'figure2.pdf', dpi=300, bbox_inches='tight')
plt.show()
```

### Gradient Model

```{r}
main_df <- read.csv('/tmp/main_df.csv')

model <- glmer(predictShared ~ experiment_num * stance_num * category_num +
               (1|pid) + (1|matchedDomain),
               data = main_df, family = binomial)
summary(model)
confint(model, method = "Wald")
```

### SR-G Model

```{r}
srg_df <- read.csv('/tmp/srg_df.csv')
model <- lmer(srgiResponse ~ predictShared * experiment_num * stance_num + (1|matchedDomain),
              data = srg_df)
summary(model)
confint(model, method = "Wald")
```

## 4. Domain Analysis (Figure 3 + Domain-Specific Models)

Paper reports: 42 models (7 domains × 3 types × 2 stances), >95% positive coefficients, all significant effects positive after FDR (p.9)

```{python}
#| fig-cap: "Figure 3: Domain-level heatmaps"
#| fig-width: 10
#| fig-height: 10
# Reload modules and data
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns
from pathlib import Path
base_dir = Path.cwd().resolve()
while base_dir.name and not (base_dir / 'data').exists():
    base_dir = base_dir.parent
figures_dir = base_dir / 'outputs' / 'figures'
# Stance palette: opposing (blue), shared (pink)
stance_palette = ['#648FFF', '#DC267F']
SAVE_FIGURES = True  # Save figures when running
df_pd = pd.read_csv('/tmp/main_df.csv', low_memory=False)

domain_order = ['Lifestyle', 'Background', 'Identity', 'Morality', 'Politics', 'Preferences', 'Religion']
category_order = ['sameQ', 'sameD', 'diffD']
category_labels = ['Same\nQuestion', 'Same\nDomain', 'Different\nDomain']

df_pd['domain_cap'] = df_pd['matchedDomain'].str.replace('arbitrary', 'Lifestyle').str.capitalize()
heatmap_data = df_pd.groupby(['experiment', 'stance', 'domain_cap', 'category'])['predictShared'].mean().reset_index()

fig, axes = plt.subplots(3, 2, figsize=(10, 10), sharex=True)

for row, exp in enumerate(['chat', 'no-chat']):
    for col_idx, stance in enumerate(['opposing', 'shared']):
        ax = axes[row, col_idx]
        subset = heatmap_data[(heatmap_data['experiment'] == exp) & (heatmap_data['stance'] == stance)]
        pivot = subset.pivot(index='domain_cap', columns='category', values='predictShared')
        pivot = pivot.reindex(index=domain_order, columns=category_order)
        sns.heatmap(pivot, ax=ax, cmap='YlOrRd', vmin=0, vmax=1, cbar=(col_idx == 1),
                    cbar_kws={'label': '% Commonality'} if col_idx == 1 else {})
        ax.set_xticklabels(category_labels, rotation=45, ha='right')
        ax.set_ylabel('Domain' if col_idx == 0 else '')
        ax.set_xlabel('')
        if row == 0:
            ax.set_title(stance.capitalize(), fontweight='bold')

# Contrast row
for col_idx, stance in enumerate(['opposing', 'shared']):
    ax = axes[2, col_idx]
    chat = heatmap_data[(heatmap_data['experiment'] == 'chat') & (heatmap_data['stance'] == stance)]
    nochat = heatmap_data[(heatmap_data['experiment'] == 'no-chat') & (heatmap_data['stance'] == stance)]
    merged = chat.merge(nochat, on=['domain_cap', 'category'], suffixes=('_chat', '_nochat'))
    merged['diff'] = merged['predictShared_chat'] - merged['predictShared_nochat']
    pivot = merged.pivot(index='domain_cap', columns='category', values='diff')
    pivot = pivot.reindex(index=domain_order, columns=category_order)
    sns.heatmap(pivot, ax=ax, cmap='RdBu_r', center=0, vmin=-0.3, vmax=0.3, cbar=(col_idx == 1),
                cbar_kws={'label': 'Chat - No-Chat'} if col_idx == 1 else {})
    ax.set_xticklabels(category_labels, rotation=45, ha='right')
    ax.set_ylabel('Domain' if col_idx == 0 else '')

axes[0, 0].text(-0.3, 0.5, 'Chat', transform=axes[0, 0].transAxes, fontweight='bold', va='center', rotation=90)
axes[1, 0].text(-0.3, 0.5, 'No-Chat', transform=axes[1, 0].transAxes, fontweight='bold', va='center', rotation=90)
axes[2, 0].text(-0.3, 0.5, 'Contrast', transform=axes[2, 0].transAxes, fontweight='bold', va='center', rotation=90)

plt.tight_layout()
if SAVE_FIGURES:
    plt.savefig(figures_dir / 'figure3_heatmaps.pdf', dpi=300, bbox_inches='tight')
plt.show()
```

### 42 Domain-Specific Models

```{r}
domain_df <- read.csv('/tmp/main_df.csv')
domain_df$experiment <- factor(domain_df$experiment, levels = c('no-chat', 'chat'))
contrasts(domain_df$experiment) <- cbind(c(-0.5, 0.5))

fit_model <- function(data, cat) {
    if (cat == 'sameQ') {
        m <- glm(predictShared ~ experiment, data = data, family = binomial())
    } else {
        m <- glmer(predictShared ~ experiment + (1|pid), data = data, family = binomial(),
                   control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 100000)))
    }
    tibble(beta = coef(summary(m))[2, 1], pVal = coef(summary(m))[2, 4])
}

results <- domain_df %>%
    group_by(matchedDomain, category, stance) %>%
    nest() %>%
    mutate(res = map2(data, category, fit_model)) %>%
    select(-data) %>%
    unnest(res) %>%
    ungroup() %>%
    mutate(pVal_fdr = p.adjust(pVal, method = 'fdr'))

n_pos <- sum(results$beta > 0)
n_sig_pos <- sum(results$pVal_fdr < 0.05 & results$beta > 0)
n_sig_neg <- sum(results$pVal_fdr < 0.05 & results$beta < 0)

cat(sprintf('Total models: %d\n', nrow(results)))
cat(sprintf('Positive coefficients: %d (%.1f%%)\n', n_pos, 100 * n_pos / nrow(results)))
cat(sprintf('Significant after FDR: %d positive, %d negative\n', n_sig_pos, n_sig_neg))
cat('\nPaper reports: >95% positive, all significant effects positive\n')
```

## 5. Sample Description

```{python}
# Reload data
import pandas as pd
df_pd = pd.read_csv('/tmp/main_df.csv', low_memory=False)

print('=== Sample Description ===')
print(f"Total participants: N = {df_pd['pid'].nunique()}")
print(f"  No-chat: n = {df_pd[df_pd['experiment'] == 'no-chat']['pid'].nunique()}")
print(f"  Chat: n = {df_pd[df_pd['experiment'] == 'chat']['pid'].nunique()}")
print(f"  Chat dyads: {df_pd[df_pd['experiment'] == 'chat']['dyadId'].nunique()}")
print(f'\nPaper reports: N = 1,205 (527 imagined; 678 real, 339 dyads)')
```
